---
title: "Iris PCA analysis"
author: "Mehmet"
date: "2024-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

## Short application of PCA to the famoous Iris data set:

```{r}
    iris_pc <- prcomp(iris[, 1:4])
    library(ggplot2)
    ggplot(data = data.frame(iris_pc$x), aes(x=PC1, y=PC2)) + 
        geom_point(color=c("chartreuse4", "darkorange", "deepskyblue")[iris$Species], size = 3)

```
One may be interested in plotting all pair-wise direction to see if lower PC’s provide useful information.

```{r}
pairs(iris_pc$x, col=c("chartreuse4", "darkorange", "deepskyblue")[iris$Species], pch = 19)
```

However, usually, the lower PC’s are less informative. This can also be speculated from the eigenvalue plot, which shows how influential each PC is.

```{r}
plot(iris_pc, type = "l", pch = 19, main = "Iris PCA Variance")
```

Feature contributions to the PC can be accessed through the magnitude of the loadings. This table shows that Petal.Length is the most influential variable on the first PC, with loading ≈0.8567.

```{r}
iris_pc$rotation
```

We can further visualize this on a plot. This can be helpful when the number of variables is large.

```{r}
features = row.names(iris_pc$rotation)
ggplot(data = data.frame(iris_pc$rotation), aes(x=PC1, y=PC2, label=features,color=features)) + 
        geom_point(size = 3) + geom_text(size=3)
```